#Intro to R 
## Before we start

* Start RStudio & show console, scripts, environments, plots
* Code & workflow more reproducible if we document
* Goal no just do stuff but to do it so can replicate results


### r as calculator

```{r}
2+5
10/5
```
To run code from script editor ctr+enter cmd+enter

How to comment & assign objects

```{r}
# this is a comment
# this is the second line
# assignment operator <-
```

```{r}
b <- 25 # = works but don't use

```

* Alt+- keyboard shortcut <- 
* functions are canned scripts
* predefined functions - typical one sqrt()
```{r}

sqrt(b)
c <- sqrt(b)
c
```


```{r}
round(3.14159)

```

* Functions have arguments, some are called 'options':
* Arguements can be defaults. 

```{r}
#use args() to find out args
args(round)
round(3.14159, digits = 2)
```

## getting help 
```{r}
?barplot
args(lm) #shows args
??geom_point # to get to package enabled functions 
help.search("kruskal") #search for fuction must be installed 
```

* Look in stackoverflow http://stackoverflow.com/questions/tagged/r

## Intro to R 

* Explain script image. function, assignment operator, = for args, ## comments, $ operator, indentation

Creating objects in console, but lost when we close session. 

```{r}
3 + 5

```

More useful to assign values to objects. 


```{r}
weight_kg <- 55 # object names sush as x, current_temp, subject_id, but cannot start with number, case sen
2.2 * weight_kg # can use object to convert to lbs
weight_kg <- 57.5 # can assign new value to the object
2.2 * weight_kg # shows different results
```

```{r}
weight_lb <- 2.2 * weight_kg #can create new object from mult
weight_lb
weight_lb <- 100  # assign new value
weight_lb
```


```{r}
#challenge
mass <- 47.5
age <- 122
(mass <- mass * 2.0)
(age <- age - 20)
(mass_index <- mass/age)
```

## Creating functions

* define a function, return value from function

```{r}
fahr_to_kelvin <- function(temp){
  kelvin <- ((temp - 32) * (5/9)) +273.15
  return(kelvin)
}
```

```{r}
# using the function
fahr_to_kelvin(32)
fahr_to_kelvin(212)
```

```{r}
kelvin_to_celsius <- function(temp){
  celsius <- temp - 273.15
  return(celsius)
}

#absolute zero in celsius 
kelvin_to_celsius(0)
```

## lets convert fahrenheit to celsius 
```{r}
fahr_to_celsius <- function(temp) {
  temp_k <- fahr_to_kelvin(temp)
  result <- kelvin_to_celsius(temp_k)
  return(result)
}

#freezing point of water in celsius
fahr_to_celsius(32.0)
```


## Vectors & data types 

* vectors the workhorse of R 
* basic data structure 

```{r}
weights <- c(50, 60, 65, 82)
weights

```
```{r}
animals <- c("mouse", "rat", "dog")
animals
```
* functions that allow you to inspect a vector. 

```{r}
length(weights)
length(animals)

```

```{r}
class(weights)
class(animals)
```

```{r}
str(weights)
str(animals)
```

* the c() function - c is for combine - creates a vector.

```{r}
#combine
weights <- c(weights, 90)
weights <- c(30, weights)
weights
```

* we just saw 2 of the 6 data types: character and numeric. Other 4 are: logical, integer, complex, raw
* vectors are one of many data structures in R: list, matrices, dataframe, factors



## Starting with data

* Let's load our survey data. 

```{r}
download.file("http://files.figshare.com/2236372/combined.csv", "data/portal_data_joined.csv")
surveys <- read.csv("data/portal_data_joined.csv")
```

* wrapping an assignment in parentheses will perform the assignment and display it at the same time.

```{r eval=FALSE}
(surveys <- read.csv("data/portal_data_joined.csv"))

```


* use head to just see six rows

```{r}

head(surveys)

```

* look at the structure of the data

```{r}

str(surveys)

```


* challenge - slide
* What is the class of object survey

## factors 

* used to represent categorical variables 
* Factors are stored as integers
* can only contain a pre-defined set values, known as levels
* alpha org 

```{r}
sex <- factor(c("male", "female", "female", "male"))
levels(sex)
nlevels(sex)
```

* sometimes order matters in factors (low med high) 


```{r}
food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(food)
food <- factor(food, levels = c("low", "medium", "high"))
levels(food)
#min(food)  ## doesn't work
```


```{r}
# ordered true
food <- factor(food, levels = c("low", "medium", "high"), ordered = TRUE)
levels(food)
min(food)  ## works!
```


### converting factors 

* use as.character(x)
* converting to numeric tricky


```{r}

f <- factor(c(1, 5, 10, 2))
as.numeric(f)  ## wrong! and there is no warning...
as.numeric(as.character(f))  ## works...
as.numeric(levels(f))[f]  ## The recommended way.

```



```{r}
## Challenge slide

## How can you recreate this plot but by having 'control' being listed last
## instead of first?
exprmt <- factor(c("treat1", "treat2", "treat1", "treat3", "treat1", "control", 
                   "treat1", "treat2", "treat3"))
table(exprmt)

barplot(table(exprmt))
```
```{r}
##anser
exprmt <- factor(exprmt, levels = c("treat1", "treat2", "treat3", "control"))
barplot(table(exprmt))
```


## Dataframes

* data.frame is the de facto data structure for most tabular data and what we use for statistics and plotting.
* created by functions read.csv() or read.table() - importing spreadsheets from file
* by default data.frame converst columns that contain characters (text) into factor data type. this might not be what you want. 
* to keep as character vectors use stringsAsFactors = False option in read.csv or read.table
* make aware of many other 'reads'

```{r}

#some_data <- read.csv("data/some_file.csv", stringsAsFactors=FALSE)

```

* you can create data.frame manually wiht data.frame() - also takes stringsAsFactors arg
* compare 

```{r}
example_data <- data.frame(animal=c("dog", "cat", "sea cucumber", "sea urchin"),
                           feel=c("furry", "furry", "squishy", "spiny"),
                           weight=c(45, 8, 1.1, 0.8))
str(example_data)
```

```{r}

example_data <- data.frame(animal=c("dog", "cat", "sea cucumber", "sea urchin"),
                           feel=c("furry", "furry", "squishy", "spiny"),
                           weight=c(45, 8, 1.1, 0.8), stringsAsFactors=FALSE)
str(example_data)
```



##challenge df - slides 

## inspecting data frames 

* size dim() nrows() ncol()
* content head() tail()
* names() rownames()
* sum str() summary()

```{r}

# Indexing 
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
#3?
```

```{r}
#using combine
animals[c(3,2)]

```

```{r}
#slice
animals[2:4]

```


```{r}
more_animals <- animals[c(1:3, 2:4)]
more_animals
```

* indexes start at 1
* `:` is a special function that creates numeric vectors of integers in increasing or decreasing order
* `seq()` use args(seq())

```{r}

seq(1, 10, by=2)

```

```{r}
# desired length of sequence
seq(5, 10, length.out=3)
```

```{r}
seq(50, by=5, length.out=10)

```

```{r}

seq(1, 8, by=3) # sequence stops to stay below upper limit

```

```{r}
#dataframe 
surveys[1, 1]   # first element in the first column of the data frame
surveys[1, 6]   # first element in the 6th column
surveys[1:3, 7] # first three elements in the 7th column
surveys[3, ]    # the 3rd element for all columns
surveys[, 8]    # the entire 8th column
head_surveys <- surveys[1:6, ] # surveys[1:6, ] is equivalent to head(surveys)

```



#Challenge
#The function nrow() on a data.frame returns the number of rows. Use it, in conjuction with seq() 
#to create a new data.frame called surveys_by_10 that includes every 10th row of the survey data 
#frame starting at row 10 (10, 20, 30, …)

surveys_by_10 <- surveys[seq(10, nrow(surveys), by=10), ]
